%Requires the memoir class 
\documentclass[twoside,11pt]{memoir}

%\usepackage{mathptmx}  % Times New Roman, but if you have Garamond 
                        % then use it;
                        % you are writing a book, not a newspaper column

\DoubleSpacing        % memoir's double spacing
\usepackage{rice}     % rice thesis package 


\usepackage{graphicx}
\usepackage{rotating}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{multicol}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{bbm}
%\usepackage{siunitx}
%\usepackage{txfonts}  % I used this one to summon the symbol for \lambdabar
\usepackage{lmodern}

\usepackage[numbers,compress]{natbib}

\usepackage[backref=page]{hyperref}
%\hypersetup{colorlinks=false}

\usepackage{parskip}
\usepackage{xfrac}
\usepackage{isomath}

\usepackage{memhfixc}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% Definition of command shortcuts                           %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\twos}[1]
     {\ensuremath{\hspace{-1pt}2\hspace{0.5pt}\hspace{0pt}S_{#1}\hspace{-1pt}}}
\newcommand{\twop}[1]
     {\ensuremath{\hspace{-1pt}2\hspace{0.5pt}\hspace{0pt}P_{#1}\hspace{-1pt}}}
\newcommand{\trep}[1]
     {\ensuremath{\hspace{-1pt}3\hspace{0.5pt}\hspace{0pt}P_{#1}\hspace{-1pt}}}

\newcommand{\cm}{\ensuremath{,\hspace{1pt}}}
\newcommand{\f}[1]{\ensuremath{F\hspace{-5pt}=\hspace{-4pt}#1}}
\newcommand{\mf}[1]{\ensuremath{m_{F}\hspace{-4pt}=\hspace{-4pt}#1}}
\newcommand{\mj}[1]{\ensuremath{m_{J}\hspace{-5pt}=\hspace{-4pt}#1}}
\newcommand{\mcm}{\ensuremath{\hspace{-4pt}}}

\newcommand{\red}
{\ensuremath{ \twos{1/2}\hspace{-0.0pt}\rightarrow\hspace{-0.0pt}\twop{3/2} }\ }
\newcommand{\uv}
{\ensuremath{ \twos{1/2}\hspace{-0.0pt}\rightarrow\hspace{-0.0pt}\trep{3/2} }\ }

\newcommand{\TD}{\ensuremath{ T_{D} }}
\newcommand{\TR}{\ensuremath{ T_{R} }}
\newcommand{\TF}{\ensuremath{ T_{F} }}

\newcommand{\one}{\ensuremath{|1\rangle }\ }
\newcommand{\two}{\ensuremath{|2\rangle }\ }

\newcommand{\isat}{ \ensuremath{ I_{\mathrm{sat}} } } 
\newcommand{\isatred}{ \ensuremath{ I_{sat\text{(red)}} } } 
\newcommand{\isatuv}{ \ensuremath{ I_{sat\text{(uv)}} } } 
\newcommand{\li} {\ensuremath{^{6}}Li\ }
\newcommand{\kb} { \ensuremath{k_{\mathrm{B}}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\bv}[1]{\ensuremath{\bm{#1}}}
\newcommand{\vo}{\ensuremath{V_{0}}}
\newcommand{\vvo}{\ensuremath{v_{0}}}
\newcommand{\bvo}{\ensuremath{\bv{V}_{0}}}
\newcommand{\er}{\ensuremath{E_{r}}}
\newcommand{\Lc}{\ensuremath{L_{\mathrm{c}}}}
\newcommand{\dsig}[1]{\ensuremath{ \frac{ d\,\sigma_{#1} }{d\,\Omega} }}
\newcommand{\dbl}{\ensuremath{ \uparrow\! \downarrow \, }}
\newcommand{\spup}{\ensuremath{ \uparrow }}
\newcommand{\spdn}{\ensuremath{ \downarrow}}


\newcommand{\ndbl} { \makebox[1em][c]{$\uparrow\! \downarrow \, $}}
\newcommand{\nvac} { \makebox[1em][c]{$0 $}}
\newcommand{\nspup}{ \makebox[1em][c]{$\uparrow$}}
\newcommand{\nspdn}{ \makebox[1em][c]{$\downarrow$}}
\newcommand{\nSspup}{ \makebox[0.7em][c]{$\uparrow$}}
\newcommand{\nSspdn}{ \makebox[0.7em][c]{$\downarrow$}}


\newcommand{\tbtwo}[4]{\scalebox{0.75}{\arraycolsep=1.0pt\def\arraystretch{0.7}$\left|\begin{array}{c|c}
\makebox[1em][c]{$#1$} &\makebox[1em][c]{$#2$} \\ \hline \makebox[1em][c]{$#3$}
&\makebox[1em][c]{$#4$} \\ \end{array}\right\rangle$ }}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Compensated lattice
\newcommand{\rdiag}{\ensuremath{ r_{\text{\tiny{111}}} } }
\newcommand{\awaist}{\ensuremath{ \alpha_{w} }}  
\newcommand{\awaistevap}{\ensuremath{ \alpha_{w,\text{evap}} }}  
\newenvironment{myblock}{%
   \list{}{\rightmargin0pt}\item\relax
   % and maybe do more stuff
}{\endlist}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendix on trap potentials 
\newcommand{\pin}{\ensuremath{ P_{\text{i}}} }
\newcommand{\pret}{\ensuremath{ P_{\text{r}}} }
\newcommand{\win}{\ensuremath{ w_{\text{in}}} }
\newcommand{\wret}{\ensuremath{ w_{\text{r}}} }
\newcommand{\wir}{\ensuremath{ w_{\text{ir}}} }

\newcommand{\pgr}{\ensuremath{ P_{\text{gr}}} }
\newcommand{\wgr}{\ensuremath{ w_{\text{gr}}} }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Phase-contrast
\newcommand{\ts}[1]{\ensuremath{\tensorsym{#1}}}

\newcommand{\efield}{\ensuremath{\bv{\mathcal{E}}}}
\newcommand{\efieldo}{\ensuremath{\mathcal{E}_{0}}}
\newcommand{\epspol}{\ensuremath{\hat{\bv{\varepsilon}}}}
\newcommand{\dpol}{\ensuremath{\bv{\mathrm{P}}}}
\newcommand{\phcChi}{\ensuremath{\zeta}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Fermi Thermometry
\newcommand{\rf}[1]{\ensuremath{ R_{\mathrm{F}_{#1} } }}
\newcommand{\ldb}{\ensuremath{ \lambda_{\mathrm{th}} }}
\newcommand{\pli}{\ensuremath{ \mathrm{Li} }}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bragg
\newcommand{\sumOff}{\ensuremath{\sum_{\genfrac{}{}{-2pt}{}{m,n}{m\neq n}}}}
\newcommand{\sumOffn}{\ensuremath{\sum_{\genfrac{}{}{-2pt}{}{n}{n\neq m}}}}
\newcommand{\sumOffm}{\ensuremath{\sum_{\genfrac{}{}{-2pt}{}{m}{m\neq n}}}}
\newcommand{\sumOffij}{\ensuremath{\sum_{\genfrac{}{}{-2pt}{}{i,j}{i\neq j}}}}
\newcommand{\scrys}{\ensuremath{S_{\bv{Q}}^{\mathrm{crystal}}}}
\newcommand{\sq}{\ensuremath{S_{\bv{Q}}}}
\newcommand{\sPi}{\ensuremath{S_{\bv{\pi}}}}
\newcommand{\zoz}{\ensuremath{(0\hspace{0.3em}1\hspace{0.3em}0)}}
\newcommand{\ijk}{\ensuremath{(i\hspace{0.3em}j\hspace{0.3em}k)}}
%\newcommand{\zoz}{\ensuremath{(0\hspace{0.3em}\protect\raisebox{0.12em}{\protect\scalebox{0.75}[0.75]{-}}\hspace{-0.02em}1\hspace{0.3em}0)}}
\newcommand{\hhh}{\ensuremath{(\protect\raisebox{0.26em}{\protect\scalebox{0.75}[0.75]{-}}\sfrac{1}{2}\ \protect\raisebox{0.26em}{\protect\scalebox{0.75}[0.75]{-}}\sfrac{1}{2}\ \protect\raisebox{0.29em}{\protect\scalebox{0.5}[0.5]{+}}\sfrac{1}{2})}}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bragg paper
\newcommand{\half}{\frac{1}{2}}
\newcommand{\expect}[1]{\ensuremath{\langle{#1}\rangle}}
\newcommand{\ket}[1]{\ensuremath{|{#1}\rangle}}
\newcommand{\mathbfsf}[1]{\ensuremath{\bm{#1}}}
\newcommand{\debyew}{\ensuremath{e^{-2W_{\bv{Q}}(\tau)}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\includeonly{
%part07,
%part08,
%part09,
%partB,
%partC,
%}

\DoubleSpacing

\begin{document}

\maxtocdepth{subsection}   % put everything into the ToC
\pagestyle{plain}          % pagestyle for the prelims

\frontmatter
\thetitlepage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% put your abstract here

\riceabstract
\pagestyle{empty}  % Rice requires no page numbering in the abstract


The Hubbard model contains only the essential ingredients to describe the
behavior of strongly interacting electrons moving in a periodic lattice.   It
describes particles that can tunnel between sites in the lattice and that
acquire an on-site interaction energy when two of them occupy the same lattice
site.   This simple model is a prominent example of how strongly correlated
phases emerge from simple Hamiltonians.  It gives rise to a Mott-metal
insulator transition,  and at a density of one particle per site shows an
antiferromagnetic ground state.  It is also considered to contain the essence
of high-temperature superconductivity as observed in the cuprates, a question
that remains open due to the difficulty in numerically accessing the solutions
of the model at densities different from one particle per site.  

In this work we have realized the Hubbard model with a spin mixture of
ultracold atoms in a simple cubic optical lattice.  Atoms in lattices have
emerged in the last decade as promising systems in which to perform quantum
simulations of condensed matter Hamiltonians.   In the laboratory we can create
defect-free optical lattice potentials with laser light, and we can control the
interactions between the atoms using a magnetic Feshbach resonance.  

For this work we implemented a novel compensated optical lattice setup, which
allows us to control the density of the sample and mitigate the
non-adiabaticity in the lattice loading process which often leads to heating of
the sample or to out of equilibrium distributions.   Using the compensated
optical lattice we are able to get closer to the ground state of the Hubbard
model than anybody before us has been able to do with ultracold atoms.

To demonstrate this achievement we use spin-sensitive Bragg scattering of light
to measure the spin-structure factor, a measure of the antiferromagnetic
correlations in the collection of spins.  Measurements of the spin-structure
factor are compared to results of theoretical calculations to establish precise
thermometry for the atoms in the lattice.  We have also studied the
\textit{in-situ} density distribution of the system, which confirms that the
temperature of our sample is in a regime where most of the remaining entropy in
the system resides in the spin degrees of freedom.  

The results presented here represent an important step in the field of quantum
simulation using ultracold atoms.  In the future, we expect to further explore
and exploit the experimental possibilities opened up by the compensated lattice
potential and by light scattering thermometry, with the ultimate goal of
addressing the existence of $d$-wave superfluidity in the Hubbard model. 


\pagestyle{plain} % Restore page numbering.

% put your acknowledgments here 
 
%\riceacknowledgments

%Acknowledgments go here. 

%% \setdedication{ text } % if you want a dedication
%\ricededication

%\tableofcontents
\tableofcontents*  % The starred version does not add "Table of Contents" 
                   % to the Table of Contents
                   % I prefer it this way


% I don't find these two particularly useful
%% \listoffigures  % if you want to include a list of figures  
%% \listoftables   % if you want to include a list of tables


%% if you have more prelim sections, then
%%%% \clearpage
%%%% \pagestyle{plain}
%%%% \prelimtitle   text % for sections after the ToC, etc, before main text


\mainmatter
\pagestyle{rice}


%% Change the spacing between paragraphs, I prefered this for readability 
\let\oldparskip\parskip
\setlength{\parskip}{0.8em}



\include{part01}
\include{part02}
\include{part03}
\include{part04}
\include{part05}
\include{part06}
\include{part07}
\include{part08}
\include{part09}










\chapter{Conclusion} 

I feel very proud to be the first student that will receive a Ph.D. for work
done in this apparatus.   We started out in 2007 and have always battled more
that what would seem necessary to get the experiment up and running.  Luckily,
we  were able to change some problems into opportunities, and now the results
of our efforts have paid off.  

The low atom numbers that we would achieve in the ODT when loading it directly
from the red MOT, led us to focus our efforts on the UVMOT, which was the
subject of my Master's thesis.   The UVMOT is basically what gives this
experiments its edge;  it works much better than anybody ever expected (except
perhaps Randy).  The 323~nm power necessary to operate it is not too high, and
it barely needs any repump light.  The UVMOT allows us to load a significant
number of atoms ($10^{7}$) into an ODT with a very moderate depth (300~$\mu$K,
moderate for lithium standards) which is sufficient for the experiments that we
carry out.

It was only until my fifth year in grad school (Fall 2011) that we started
building up the compensated lattice.   We first observed Bragg scattering from
the \zoz\ planes in the summer of 2012.   That moment was an important turning
point, from then on most of our time was no longer spent building something new
for the apparatus but actually running experiments. 

The last two years have been almost fully dedicated to achieving our goal of
measuring the spin structure factor in the lattice using Bragg scattering of
light.   Somehow everybody thought that the Bragg signal would be like a
beacon, but we soon realized that we were looking for a needle in a haystack.
We were forced to rely heavily on statistics to bring the AFM correlations
signal out of the background.     The complications with the signal to noise
ratio led to my efforts in understanding the Bragg scattering measurement in
the limit where the signal is just above the incoherent scattering background.
Furthermore, in order to consolidate our results, we saw the need to collaborate
closely with theorists.  I quickly realized how little I new about the Hubbard
model which I had been trying to realize experimentally, but I saw here another
opportunity to serve as the bridge between the theoretical results and our
experimental measurements.

Understanding the theoretical results available for the Hubbard model, and
implementing the local density approximation  has led to the possibility of
using Bragg scattering as a thermometer for atoms in a lattice.  This is a
technique that we hope will be useful for other groups.  In our apparatus we
hope to continue perfecting the measurement of AFM correlations and use it as a
guiding light to find ways of squeezing more entropy out of the system and
bring it closer to realizing the full potential of quantum simulation.
Studying pairing may not be so far after all, given that some entropy can be
squeezed out of the system by increasing the lattice depth along one of the
axis, which would create a set of uncorrelated two-dimensional (2D) planes,
each one an independent realization of the 2D Hubbard model.  As was already
shown for 1D chains~\cite{Greif2013}, when you have a fixed amount of entropy
in the total system you can ``divide and conquer''.   The big challenge then
would be figuring out how to obtain information about the 2D systems in our
setup, given that Bragg scattering as presented here, would not work between
the uncorrelated planes.   Nevertheless, the 3D system can be used in that case
solely for the purpose of thermometry.  

In the near future, our experiment will undergo a significant upgrade to a new
compensated optical lattice setup.   The setup has been carefully designed to
implement the ideas that were presented in
Chapter~\ref{chap:compensated-optical-lattice} of this thesis.   It also boasts
superior mechanical stability, which we hope will eliminate the need to realign
the compensated lattice every day. 

In the near future, there are also plans to explore physics in one dimension
(1D) using our apparatus. This geometry can be created in our system by simply
turning off one of the lattice beams, giving rise to an array of
one-dimensional (1D) tubes.   The physics of 1D systems provides an interesting
test-bed for quantum simulation, since exact solutions exist for the many-body
problem in 1D.  

I believe that it is a good idea for the apparatus to diversify and try to study
systems for which the temperature requirements are not as low as in the
strongly correlated regime of the 3D Hubbard model.   The 1D systems mentioned
above are one example,  but also there are a lot of interesting possibilities
such as studying the Hubbard model with attractive interactions or also trying
to develop techniques to study the dynamics of systems out of equilibrium.   

The apparatus has finally reached maturity. It is a versatile machine, capable
of producing some the coldest degenerate Fermi gases of ultracold atoms ever
recorded.  I hope that the new generations of graduate students that come to it
have as much fun as I did trying to figure things out.  


%% finally, start of your main text

% if appendices, then
\appendix
\include{partA} 
\include{partB}
%\include{partC}

% if Biographical sketch then
%\begin{biosketch} I was kissed by a donkey when I was 5..... and text and more
%text and even more text, lots of text, here and there, more and more text,
%with more and more descriptions of text, blah etc blah et \end{biosketch}

\bibliographystyle{osa}
% Other options for the bibliographystyle are listed below
%\bibliographystyle{unsrt}
%\bibliographystyle{unsrtnat}
%\bibliographystyle{apsrev}
\bibliography{pmd}

\end{document}

